Unveiling Toronto's Respiratory Landscape

### 2024-01-25 outputs/paper/paper.qmd
Update paper.qmd

---
title: "Unveiling Toronto's Respiratory Landscape"
subtitle: "Analyzing Post-COVID-19 Infection Trajectory and Rhinovirus Surges through 2020 and 2023 Datasets"
author: 
  - Hailey Jang
thanks: "Code and data supporting this analysis is available at: https://open.toronto.ca/dataset/outbreaks-in-toronto-healthcare-institutions/."
date: January 23, 2024
date-format: long
abstract: "This research investigates the impact of COVID-19 on respiratory infections in Toronto, utilizing datasets from 2020 and 2023. Through analysis of simulated data on respiratory infection rates, with a focus on Rhinovirus, a noticeable surge in respiratory ailments post-pandemic is revealed. Statistical examinations comparing infection rates offer insights into the changing dynamics of respiratory health in the city. This study contributes valuable information towards understanding the lasting effects of COVID-19 on respiratory well-being in the Toronto population, providing a crucial foundation for public health strategies and interventions."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Data 
All the data used in this study were sourced form the Toronto Open Data portal [@torontodata].The dataset, titled 'Outbreaks in Toronto Healthcare Institutions' serves as the foundation for our analyses.

## Respiratory Infections in 2020
The examination of respiratory outbreak reports in Toronto healthcare institutions for the years 2020 and 2023 unveils intriguing trends that provide valuable insights into the dynamic nature of infectious diseases within these settings. Graph 1, representing 2020, showcases a consistent theme where respiratory infections notably outpace Enteric illnesses, reflecting the persistent challenges posed by respiratory health issues. The prevalence of respiratory outbreaks in this year aligns with historical expectations.

```{r}
#| echo: false
#| include: true
#| message: false
#| label: Sample of Data
#| tbl-cap: Reports of Respiratory Outbreaks (2020)

@@

```
## Respiratory Infections in 2023
Contrastingly, Graph 2, representing the post-COVID period in 2023, tells a more intricate story. While the baseline trend of respiratory infections surpassing Enteric illnesses persists, there is a marked escalation in the frequency of respiratory outbreaks compared to the preceding year. This notable surge in respiratory infection rates may be attributed to several factors, including the lingering effects of the COVID-19 pandemic. The classification of 2023 as post-COVID suggests a potential connection between the ongoing impact of the pandemic and the heightened prevalence of respiratory infections.
```{r}
#| echo: false
#| include: true
#| message: false
#| label: Sample of Data 2023
#| tbl-cap: Reports of Respiratory Outbreaks (2023)

@@

# Plotting
ggplot(filtered_data_2023, aes(x = `Type of Outbreak`, fill = `Type of Outbreak`)) +
  geom_bar() +
  labs(title = "Reports of Respiratory Outbreaks in Toronto Healthcare Institutions (2023)",
       x = "Type of Outbreak",
       y = "Count") +
  theme_minimal()
```

@@

The graphical representation (graph 3) comparing the rates of causative agents for respiratory outbreaks in Toronto healthcare institutions in 2020 and 2023 provides insightful observations. Notably, the data underscores a significant shift in the prominence of Rhinovirus as a causative agent between the two years. In 2023, Rhinovirus emerges as the predominant viral contributor, reporting a notably higher incidence compared to its occurrence in 2020.

@@

```{r}
#| echo: false
#| include: true
#| message: false
#| label: Sample of Causative Agents
#| tbl-cap: Rates of Causative Agents (2020 vs. 2023)

@@

# Plotting
ggplot(combined_data, aes(x = Causative_Agent, fill = Year)) +
  geom_bar(position = "dodge") +
  labs(title = "Rates of Causative Agents (2020 vs. 2023)",
       x = "Causative Agent",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The graph highlights a substantial increase in Rhinovirus cases in 2023, positioning it as the primary causative agent for respiratory outbreaks in Toronto healthcare institutions. This observation prompts further exploration into the factors contributing to this shift and underscores the importance of ongoing surveillance and research to adapt public health strategies to the evolving landscape of respiratory infections.

@@

The distinctive impact of social measures on Rhinovirus can be attributed to its intrinsic characteristics—its small size and capsid coat may render it less susceptible to masking and disinfectants. The observed partial increase in asthma hospitalizations following school reopening suggests an intricate relationship between Rhinovirus infections and asthma exacerbations. Importantly, admissions, while experiencing an uptick, remained lower than pre-COVID levels, indicating a nuanced interplay between respiratory infections, social dynamics, and asthma health outcomes.

@@

>>>>

---
title: "Unveiling Toronto's Respiratory Landscape"
subtitle: "Analyzing Post-COVID-19 Infection Trajectory and Rhinovirus Surges through 2020 and 2023 Datasets"
author: 
  - Hailey Jang
thanks: "Code and data supporting this analysis is available at: https://open.toronto.ca/dataset/outbreaks-in-toronto-healthcare-institutions/."
date: January 23, 2024
date-format: long
abstract: "This research investigates the impact of COVID-19 on respiratory infections in Toronto, utilizing datasets from 2020 and 2023. Through analysis of simulated data on respiratory infection rates, with a focus on Rhinovirus, a noticeable surge in respiratory ailments post-pandemic is revealed. Statistical examinations comparing infection rates offer insights into the changing dynamics of respiratory health in the city. This study contributes valuable information towards understanding the lasting effects of COVID-19 on respiratory well-being in the Toronto population, providing a crucial foundation for public health strategies and interventions."
output: pdf_document
number-sections: true
format: pdf
bibliography: references.bib
---

@@

# Data 
All the data used in this study were sourced form the Toronto Open Data portal [@opendatatoronto].The dataset, titled 'Outbreaks in Toronto Healthcare Institutions' serves as the foundation for our analyses.

For the purpose of data analysis, simulation, cleansing, and visualization, the statistical programming language R was judiciously employed [@citeR]. Additionally, to facilitate these analytical processes, a suite of specialized R packages were utilized, including but not limited to here [@citeggplot2] and the comprehensive tidyverse suite [@tidyverse].

## Respiratory Infections in 2020
The examination of respiratory outbreak reports in Toronto healthcare institutions for the years 2020 and 2023 unveils intriguing trends that provide valuable insights into the dynamic nature of infectious diseases within these settings. @fig-1, representing 2020, showcases a consistent theme where respiratory infections notably outpace Enteric illnesses, reflecting the persistent challenges posed by respiratory health issues. The prevalence of respiratory outbreaks in this year aligns with historical expectations.

```{r}
#| echo: false
#| eval: true
#| label: fig-1
#| fig-cap: "Reports of Respiratory Outbreaks (2020)"
#| warning: false

@@

```
\newpage
## Respiratory Infections in 2023
Contrastingly, @fig-2, representing the post-COVID period in 2023, tells a more intricate story. While the baseline trend of respiratory infections surpassing Enteric illnesses persists, there is a marked escalation in the frequency of respiratory outbreaks compared to the preceding year. This notable surge in respiratory infection rates may be attributed to several factors, including the lingering effects of the COVID-19 pandemic. The classification of 2023 as post-COVID suggests a potential connection between the ongoing impact of the pandemic and the heightened prevalence of respiratory infections.
```{r}
#| echo: false
#| eval: true
#| label: fig-2
#| fig-cap: "Reports of Respiratory Outbreaks (2023)"
#| warning: false

@@

# Plotting
ggplot(filtered_data_2023, aes(x = `Type of Outbreak`, fill = `Type of Outbreak`)) +
  geom_bar() +
  labs(title = "Reports of Respiratory Outbreaks in Toronto Healthcare Institutions (2023)",
       x = "Type of Outbreak",
       y = "Count") +
  theme_minimal()
```
##Prevalence of Respiratory Infections
The analysis of respiratory infection trends in the aftermath of the COVID-19 pandemic reveals a noticeable increase in the overall prevalence of respiratory infections throughout the year 2023, presenting a distinct contrast with the preceding year, 2020. This notable rise extends beyond pre-pandemic norms, signifying a substantial transformation in the landscape of respiratory health, as highlighted in (@Maison). The observable surge in respiratory infections post-pandemic underscores the enduring impact of the COVID-19 pandemic on the broader respiratory health ecosystem. Consequently, there is a compelling need for an in-depth exploration of potential contributing factors and the development of adaptive public health strategies to effectively address the evolving challenges within the realm of respiratory infections.

@@

The graphical representation, @fig-3, comparing the rates of causative agents for respiratory outbreaks in Toronto healthcare institutions in 2020 and 2023 provides insightful observations. Notably, the data underscores a significant shift in the prominence of Rhinovirus as a causative agent between the two years. In 2023, Rhinovirus emerges as the predominant viral contributor, reporting a notably higher incidence compared to its occurrence in 2020.

@@

```{r}
#| echo: false
#| eval: true
#| label: fig-3
#| fig-cap: "Rates of Causative Agents (2020 vs. 2023)"
#| warning: false

@@

# Plotting
ggplot(combined_data, aes(x = Causative_Agent, fill = Year)) +
  geom_bar(position = "dodge") +
  labs(x = "Causative Agent",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The @fig-3 highlights a substantial increase in Rhinovirus cases in 2023, positioning it as the primary causative agent for respiratory outbreaks in Toronto healthcare institutions. This observation prompts further exploration into the factors contributing to this shift and underscores the importance of ongoing surveillance and research to adapt public health strategies to the evolving landscape of respiratory infections.

@@

The discernible influence of social interventions on Rhinovirus underscores the significance of its intrinsic attributes, particularly its diminutive size and capsid coat. According to Kreger (@Kreger), these characteristics may contribute to its reduced susceptibility to masking agents and disinfectants. The observed incremental surge in asthma-related hospitalizations subsequent to the reopening of schools implies a complex interconnection between Rhinovirus infections and exacerbations of asthma. Significantly, notwithstanding the increase in admissions, the levels remained below those observed prior to the onset of the COVID-19 pandemic, revealing a nuanced interplay among respiratory infections, societal dynamics, and health outcomes associated with asthma.

\newpage

@@

# Reference
